### 2018-04-12 Writing/disparity_review.tex
Updated data bit

\noindent{\Large \bf Disparity in morphological disparity analysis} 
% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
%TG: back to the drawing board. Maybe something with "best practices" as well.
% NC: how about something like "Analysing disparity in morphology data: best practice and future directions"
\bigskip

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

% NC: OK I've chopped this up a lot as it felt like you were repeating yourself quite a bit (e.g. you said here is what kinds of analyses people did in the 90s, and what they do now, then you had a section on here is what we defined disparity as in the 1980s and here is what we do now - so I've combined it all) and could say the same thing in half the space. Feel free to revert back.
\section{Introduction}

% NC: New version
Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.

In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF).
Early disparity analyses focused on broad questions such as How does diversity evolve? Are some body types more common than others? Can shape evolve in all ``directions'' (of the multidimensional space)?
Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e. developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e. intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e. responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).

@@

% OLD VERSION
\subsection{old version of intro}
%Even classical scholars recognised that NC: I'm unsure we need this qualifier, just get right to it!
Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
% NC: I've tried to make the example clearer, but I'm unsure if it actually helps. Could cut the examples and go straight into the next sentence? Then the intro is just one paragraph.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as morphological disparity analysis.

This type of analysis was popularised in the 1980s and 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] through studies of the fossil record, focused on broad questions such as how does diversity evolve? Are some body types more common than others? Can shape evolve in all ``directions'' (of the multidimensional space)?

In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

Since then, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity'' as a synonym for ``analysis of patterns of morphological diversity''.
Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e. developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e. intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e. responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).

What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).

Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.
Thus, disparity methods or analysis can be seen as methods or analysis to study these patterns and eventually link them to macroevolutionary or macroecological processes

@@

\section{Collect the appropriate data}
Once you 
%TG: not sure about using "you" all the time. I feel it does add some patronising tone the whole yoke.
have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over purely evolutionary traits.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).

\subsection{Data types}
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can be either describing a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or counts (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted methods (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item or any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question. % NC: This section might need more explanation and/or some more examples - %TG: here's a try.
This illustrates the importance of framing the question.
In this above example, if the question is whether the aerodynamic properties of wings varies within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
Conversely if the question focuses on phylogenetic convergence aspects between different types of active flights, it might be more pertinent to use evolutionary traits that facilitate active flights within and between both groups (e.g. digits length, head shape, etc.).

\subsection{How much data to collect?}
Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of both specimens and traits.
There are several important considerations.
\begin{itemize}
    \item \textit{Taxonomic scope}
    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
    \item \textit{Sample size}
    There is always a trade-off between collecting enough data and the time and resources allocated to a project.
    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction analysis [CITE]).
    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
    % \item \textit{Are the data consistent?}
        %NC: Not suer I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
        %TG: fair enough. This was kind of the miscellaneous section steming from the discussion on data (bullet points about the importance of sexual dimorphism, etc...)
    % This is maybe the most important question related to the nature of the data.
    % It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
    % For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
    % Are the elements fairly consistent and comparable them?
    % Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
    % In any of these cases, are the patterns a mere reflection of these inconsistencies?
    % For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g. snakes) that could not reflect processes.
\end{itemize}
Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.

\subsection{Recycling data}
A common practice in disparity analysis is to recycle data from other studies which can introduce additional problems.
One classic case is using ``cladistic'' data in palaeobiology.
These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
This data can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all the data.
For example, in a study of disparity of dietary niches, removing non-dietary characters is a good approach.
However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g. limb traits can be used for both locomotion and feeding).

@@

\section{Use appropriate methods} 
In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
Unfortunately, we can only perceive a maximum of three dimensions at the same time.
Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions. 
In this example, no group overlaps in the multidimensional space despite the appearances.}} 
% NC: Might need to explain more clearly, I'm not 100% sure I got your meaning
\label{Fig:RF_results_best}
\end{figure}

It is therefore vital to be clear about what is represented in multidimensional space visualisations.
In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
We suggest visualisations should include the following:
(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g. disparity through time).
Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading.

\subsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
\label{box_ordination}
One of the most common approaches to analysing multidimensional data is to use ordination techniques.
These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
This is generally done via three analysis pipelines:
\begin{enumerate}
    \item data -> distance matrix -> ordination
    \item data -> zscores -> principal components analysis (PCA)
    \item data -> Procrustes superimposition -> projection into tangent space -> PCA
\end{enumerate}
%TG: Check with Melanie for the figure? - apparently she has one.

%Why to ordinate: % NC: This box seems quite long, can it be reduced?
These pipelines reduce the dimensionality of the data.
This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g. selecting only the axes that describe up to 95\% of the variation).
In the case of geometric morphometric data (point 3 above), ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc. % NC: is the stuff after the : needed in this sentence?
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
This also allows easy interpretation of the proximity of the elements in space which reflect shared biological characters.
%%%%%%%%%%%
However, some caveats can arise from such transformations of the data.
For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g. if the first few axes represent the elongation and flattening of the beak, but the question is about wing disparity).
In many cases, depending on the question, ordination might not be necessary.
For example, if the disparity metric being used (see below) relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data is non-Euclidean (e.g. discrete morphological characters), the ordination can be problematic (due to character hierarchy, inapplicable characters, polymorphisms, and missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless (e.g. the ratio of data coded as 1 rather than 0). % NC: Need to explain this point more clearly

@@

>>>>

\noindent{\Large \bf Analysing disparity in morphology data: best practice and future directions}

\bigskip

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}

Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.

In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
Early disparity analyses focused on broad questions such as: How does diversity evolve? Are some body types more common than others? Can shape evolve in all ``directions'' (of the multidimensional space)?
Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e. developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e. intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e. responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).

@@

% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).

% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).

@@

\section{Collect the appropriate data}
Once one has a clear question, it is important to collect the appropriate data to avoid some major caveats.
For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over solely evolutionary traits.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).

\subsection{Data types}
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can be either describing a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or counts (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted methods (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item or any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
This illustrates the importance of framing the question.
In this above example, if the question is whether the aerodynamic properties of wings varies within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
Conversely if the question focuses on phylogenetic convergence aspects between different types of active flights, it might be more pertinent to use evolutionary traits that would have facilitated active flights within and between both groups (e.g. digits length, integumentary system, etc.).

\subsection{How much data to collect?}
Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
\begin{itemize}
    \item \textit{Taxonomic scope}
    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
    \item \textit{Sample size}
    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction analysis [CITE]).
    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
    % \item \textit{Are the data consistent?}
        %NC: Not sure I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
        %TG: fair enough. This was kind of the miscellaneous section steming from the discussion on data (bullet points about the importance of sexual dimorphism, etc...)
    % This is maybe the most important question related to the nature of the data.
    % It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
    % For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
    % Are the elements fairly consistent and comparable them?
    % Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
    % In any of these cases, are the patterns a mere reflection of these inconsistencies?
    % For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g. snakes) that could not reflect processes.
\end{itemize}
Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.

\subsection{Recycling data}
A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
One classic case is using ``cladistic'' data in palaeobiology.
These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all the data.
For example, in a study of disparity of dietary niches, removing non-dietary characters is a good approach.
However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g. limb traits can be used for both locomotion and feeding).

@@

\section{Use the appropriate methods} 
In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
Unfortunately, we can only perceive a maximum of three dimensions at the same time.
Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions.
In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
Although these groups overlap in the two first dimensions (panel 1), they do not overlap between the dimension 1 and 3 thus do no overlap in the multidimensional space at all.
The inverse is of course true as well: if two groups don't overlap between two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap between any other combination of dimensions (dimensions 1 and 3 in panel 2).}} 
\label{Fig:RF_results_best}
\end{figure}

It is therefore vital to be clear about what is represented in multidimensional space visualisations.
In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
We suggest visualisations should include the following:
(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g. disparity through time).
Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).

\subsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
\label{box_ordination}
% One of the most common approaches to analysing multidimensional data is to use ordination techniques.
% These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
% This is generally done via three analysis pipelines:

% \begin{enumerate}
%     \item data -> distance matrix -> ordination
%     \item data -> zscores -> principal components analysis (PCA)
%     \item data -> Procrustes superimposition -> projection into tangent space -> PCA
% \end{enumerate}
% %TG: Check with Melanie for the figure? - apparently she has one.

Reducing the dimensionality through ordination of the data can be advantageous for plotting and visualising the data which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g. selecting only the axes that describe up to 95\% of the variation).
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
%the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.%TG: removed below for now.
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).

However, some caveats can arise from such transformations of the data.
For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g. if the first few axes represent the elongation and flattening of the beak, but the question is about wing disparity).
In many cases, depending on the question, ordination might not be necessary.
For example, if the disparity metric being used (see below) relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data is non-Euclidean (e.g. discrete morphological characters), the ordination can be problematic (due to character hierarchy, inapplicable characters, polymorphisms, and missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless (e.g. the ratio of data coded as 1 rather than 0). % NC: Need to explain this point more clearly

@@

%TG: add the curse of dimensionality
